---
title: "MH_MCMC"
output: html_document
---

# Load, preprocess and make design matrices

```{r}
heart_df <- read.csv("data/heart.csv")

# Center and Scale the continuous variables
heart_df$Age <- scale(heart_df$Age)
heart_df$MaxHR <- scale(heart_df$MaxHR)
heart_df$Oldpeak <- scale(heart_df$Oldpeak)

# Recode binary variables
heart_df$binary_sex <- ifelse(heart_df$Sex == "M", 1, 0)
heart_df$binary_angina <- ifelse(heart_df$ExerciseAngina == "Y", 1, 0)

# Select features for the basic model
basic_model <- model.matrix(~ Age + MaxHR + Oldpeak + binary_sex + binary_angina + ST_Slope,
                  data = heart_df)

# Select features for the more complex model (interactions + non-linearity)
complex_model <- model.matrix(~ Age*MaxHR + Oldpeak + binary_sex + binary_angina + ST_Slope + I(Age^2),
                  data = heart_df)

```

# Define log-prior, log-likelihood and log-posterior functions

```{r}
# Gaussian prior: beta ~ N(0, sigma2)
# Returns log-prior up to an additive constant (constants omitted for MH)
log_prior <- function(beta, sigma2) {
  -sum(beta^2) / (2 * sigma2)
}

# Logistic regression log-likelihood
# X %*% beta gives the linear predictor (log-odds)
# plogis() converts log-odds to probabilities
# dbinom(..., log = TRUE) computes log-likelihood in a stable way
log_likelihood <- function(beta, X, y) {
  linear_predictor <- X %*% beta
  p <- plogis(linear_predictor)
 
  sum(dbinom(y, size = 1, prob = p, log = TRUE))
}


# Log-posterior = log-likelihood + log-prior
log_posterior <- function(beta, X, y, sigma2) {
  log_likelihood(beta, X, y) + log_prior(beta, sigma2)
}
```

# Implement the Metropolis-Hastings MCMC sampler
```{r}

# One MH update step
metropolis_hastings <- function(beta_current, log_post_current, X, y, step_size, sigma2) {
 
  # propose new beta
  beta_proposal <- beta_current + rnorm(length(beta_current), 0, step_size)
 
  # evaluate posterior at proposal
  log_post_proposal <- log_posterior(beta_proposal, X, y, sigma2)
 
  # acceptance probability (log scale)
  log_alpha <- log_post_proposal - log_post_current
 
  if (log(runif(1)) < log_alpha) {
    list(
      beta = beta_proposal,
      log_post = log_post_proposal,
      accepted = 1
    )
  } else {
    list(
      beta = beta_current,
      log_post = log_post_current,
      accepted = 0
    )
  }
}


# Full MH sampler
mh_mcmc <- function(X, y, n_iter, step_size, sigma2) {
 
  n_coef <- ncol(X)
 
  beta_samples <- matrix(NA, n_iter, n_coef)
 
  # initialize
  beta_current <- rep(0, n_coef)
  log_post_current <- log_posterior(beta_current, X, y, sigma2)
 
  accept_count <- 0
 
  for (i in 1:n_iter) {
   
    step <- metropolis_hastings(
      beta_current,
      log_post_current,
      X, y,
      step_size,
      sigma2
    )
   
    # update state
    beta_current <- step$beta
    log_post_current <- step$log_post
   
    # store sample
    beta_samples[i, ] <- beta_current
   
    accept_count <- accept_count + step$accepted
  }
 
  list(
    samples = beta_samples,
    acceptance_rate = accept_count / n_iter
  )
}


# Takes an array of MCMC beta coefficient samples and drops the first `n_skip` rows,
# then uses the remaining rows to generate predictions from X for each row of betas.
# The mean prediction value for each row in X is returned.
mh_predict <- function(beta_samples, X, n_skip) {
  # Drop burn-in samples (initial iterations)
  post_samples <- beta_samples[(n_skip + 1):nrow(beta_samples), ]
 
  # Number of observations and posterior draws
  n_obs <- nrow(X)
  n_draws <- nrow(post_samples)
 
  # Store predicted probabilities for each draw
  prob_matrix <- matrix(NA, nrow = n_obs, ncol = n_draws)
 
  linear_preds <- X %*% t(post_samples)
 
  # Return posterior mean probability for each observation
  rowMeans(plogis(linear_preds))
}

```

# Evolution of the beta coefficients (trace plots)
```{r}
basic_betas <- mh_mcmc(X = basic_model,
                 y = heart_df$HeartDisease,
                 n_iter = 20000,
                 step_size = 0.28, # Is this tuneable?
                 sigma2 = 25)$samples

colnames(basic_betas) <- colnames(basic_model)
basic_betas <- as.data.frame(basic_betas)
basic_betas$iter <- 1:nrow(basic_betas)

basic_betas_long <- reshape2::melt(basic_betas, id.vars = "iter",
                variable.name = "Coefficient",
                value.name = "Value")

ggplot2::ggplot(basic_betas_long, aes(x = iter, y = Value)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~Coefficient, scales = "free") +
  labs(title = "Trace Plots for MH-MCMC Coefficients in Basic Model",
       x = "Iteration", y = "Value") +
  theme_minimal()






complex_betas <- mh_mcmc(X = complex_model,
                 y = heart_df$HeartDisease,
                 n_iter = 20000,
                 step_size = 0.1, # Is this tuneable?
                 sigma2 = 25)$samples

colnames(complex_betas) <- colnames(complex_model)
complex_betas <- as.data.frame(complex_betas)
complex_betas$iter <- 1:nrow(complex_betas)

complex_betas_long <- reshape2::melt(complex_betas, id.vars = "iter",
                variable.name = "Coefficient",
                value.name = "Value")

ggplot2::ggplot(complex_betas_long, aes(x = iter, y = Value)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~Coefficient, scales = "free") +
  labs(title = "Trace Plots for MH-MCMC Coefficients in Complex Model",
       x = "Iteration", y = "Value") +
  theme_minimal()
```

# Cross-Validation

```{r}
cv_mh_mcmc <- function(X, y, n_folds = 5,
                       n_iter = 2000,
                       step_size = 0.28,
                       sigma2 = 25,
                       n_skip = 500) {
 
  set.seed(123)
 
  fold_id <- caret::createFolds(y, k = n_folds, list = FALSE, returnTrain = FALSE)
 
  mh_oof_preds <- numeric(length(y))
 
  for (k in 1:n_folds) {
   
    # Train / test split
    X_train <- X[fold_id != k, ]
    X_test  <- X[fold_id == k, ]
    y_train <- y[fold_id != k]
    y_test  <- y[fold_id == k]
   
    # Scaling based on train data, then apply to test data (no leakage)
    for (col in c("Age", "MaxHR", "Oldpeak")) {
      mu  <- mean(X_train[, col])
      sig <- sd(X_train[, col])
      X_train[, col] <- (X_train[, col] - mu) / sig
      X_test[, col] <- (X_test[, col]  - mu) / sig
    }
   
    # Fit MH-MCMC
    fit <- mh_mcmc(
      X = X_train,
      y = y_train,
      n_iter = n_iter,
      step_size = step_size,
      sigma2 = sigma2
    )
   
    # Predict (posterior mean)
    mh_oof_preds[fold_id == k] <- mh_predict(
      beta_samples = fit$samples,
      X = X_test,
      n_skip = n_skip
    )
   
  }
 
  list(
    mh_oof_preds = mh_oof_preds
  )
}
```




# Evaluating results

Currently, this section works by running cross validation, then stacking all the out-of-fold predictions inot one vector, then comparing the out of fold predictions once against the ground truth. Another possible approach is to do per-fold comparisons, then average them together.

### Re-loading the data
```{r}
# Reload the data without scaling to prevent data leakage during CV
heart_df <- read.csv("data/heart.csv")

# Center and Scale the continuous variables
heart_df$Age <- scale(heart_df$Age)
heart_df$MaxHR <- scale(heart_df$MaxHR)
heart_df$Oldpeak <- scale(heart_df$Oldpeak)

# Recode binary variables
heart_df$binary_sex <- ifelse(heart_df$Sex == "M", 1, 0)
heart_df$binary_angina <- ifelse(heart_df$ExerciseAngina == "Y", 1, 0)

# Select features for the basic model
basic_model <- model.matrix(~ Age + MaxHR + Oldpeak + binary_sex + binary_angina + ST_Slope,
                  data = heart_df)

# Select features for the more complex model (interactions + non-linearity)
complex_model <- model.matrix(~ Age*MaxHR + Oldpeak + binary_sex + binary_angina + ST_Slope + I(Age^2),
                  data = heart_df)
```

### Run cross validation and visualize results
```{r}

basic_model_cv <- cv_mh_mcmc(basic_model, heart_df$HeartDisease)
complex_model_cv <- cv_mh_mcmc(complex_model, heart_df$HeartDisease)


brier_score <- function(y, p_hat) {
  mean((y - p_hat)^2)
}

cat("Basic model Brier score: ", brier_score(heart_df$HeartDisease, basic_model_cv$mh_oof_preds), "\n")
cat("Complex model Brier score:", brier_score(heart_df$HeartDisease, complex_model_cv$mh_oof_preds), "\n")

# Calibration data
df_cal <- rbind(
  data.frame(p_hat = basic_model_cv$mh_oof_preds,   y = heart_df$HeartDisease, model = "MH_Basic"),
  data.frame(p_hat = complex_model_cv$mh_oof_preds, y = heart_df$HeartDisease, model = "MH_Complex")
)

df_cal$bin <- cut(df_cal$p_hat, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)

calibration <- df_cal %>%
  group_by(model, bin) %>%
  summarise(
    mean_pred = mean(p_hat),
    actual    = mean(y),
    .groups   = "drop"
  )

# Plot
ggplot(calibration, aes(x = mean_pred, y = actual, color = model)) +
  geom_point(size = 3) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Calibration Plot (Basic vs Complex)",
    x     = "Predicted Probability",
    y     = "Observed Frequency"
  ) +
  theme_minimal()


```

